---
output: html_document
---
<!--
%% \VignetteEngine{knitr::rmarkdown}
%% \VignetteIndexEntry{ISB Cancer Genomics Cloud}
-->

# ISB Cancer Genomics Cloud

If you have not yet setup for Google Cloud Platform, please follow [these instructions](http://googlegenomics.readthedocs.org/en/latest/workshops/bioc-2015.html) to get setup and then come back to this vignette.

This workshop consists of two parts:

 * Part 1: Use the [bigrquery](http://cran.r-project.org/web/packages/bigrquery/index.html) package to analyze Variants via [Google BigQuery](https://cloud.google.com/bigquery/what-is-bigquery) 
 * Part 2: Analyze pearson correlation between expression and methylation data.
 * Part 3: Analyze spearman correlation between expression and protein data.
 * Part 4: Create groups based on mutation, and compare gene expression using a T-test.
 
Part 1: Analyze Variants with BigQuery
======================================

Open the code and the rendered vignette in separate windows:

* Open the vignette to run it line-by-line via:
```{r eval=FALSE}
file.edit(file.path(system.file(package = "ISBCGCExamples"), "doc", "BigQueryIntroduction.Rmd"))
```

* View the rendered vignette 
```{r eval=FALSE}
file.show(file.path(system.file(package = "ISBCGCExamples"), "doc", "BigQueryIntroduction.html"))
```

Alternatively, you can view the documentation for the [ISBCGCExamples](https://github.com/isb-cgc/examples-R) package by running `help(package="ISBCGCExamples")` and clicking on *"User guides, package vignettes and other documentation."*.

Part 2: Expression and Methylation Correlation
=======================================================

Open the code and the rendered vignette in separate windows:

* Open the vignette to run it line-by-line via:
```{r eval=FALSE}
file.edit(file.path(system.file(package = "ISBCGCExamples"), "doc", "ExpressionAndMethylationCorrelation.Rmd"))
```

* View the rendered vignette 
```{r eval=FALSE}
file.show(file.path(system.file(package = "ISBCGCExamples"), "doc", "ExpressionAndMethylationCorrelation.html"))
```

Alternatively, you can view the documentation for the [ISBCGCExamples](https://github.com/isb-cgc/examples-R) package by running `help(package="ISBCGCExamples")` and clicking on *"User guides, package vignettes and other documentation."*.

Part 3: Expression and Protein Correlation
=======================================================

Open the code and the rendered vignette in separate windows:

* Open the vignette to run it line-by-line via:
```{r eval=FALSE}
file.edit(file.path(system.file(package = "ISBCGCExamples"), "doc", "ExpressionAndProteinCorrelation.Rmd"))
```

* View the rendered vignette 
```{r eval=FALSE}
file.show(file.path(system.file(package = "ISBCGCExamples"), "doc", "ExpressionAndProteinCorrelation.html"))
```

Alternatively, you can view the documentation for the [ISBCGCExamples](https://github.com/isb-cgc/examples-R) package by running `help(package="ISBCGCExamples")` and clicking on *"User guides, package vignettes and other documentation."*.

Part 4: Creating groups and performing T-tests.
======================================================================================

In this vignette we create groups based on mutation status, and compare gene expression levels using a T-test.

Open the code and the rendered vignette in separate windows:

* Open the vignette to run it line-by-line via:
```{r eval=FALSE}
file.edit(file.path(system.file(package = "ISBCGCExamples"), "doc", "GenomicAndExpression_T_test.Rmd"))
```

* View the rendered vignette 
```{r eval=FALSE}
file.show(file.path(system.file(package = "ISBCGCExamples"), "doc", "GenomicAndExpression_T_test.html"))
```

Alternatively, you can view the documentation for the [ISBCGCExamples](https://github.com/isb-cgc/examples-R) package by running `help(package="ISBCGCExamples")` and clicking on *"User guides, package vignettes and other documentation."*.




