
# Accessing the GDC endpoints #
The new GDC website went live June 6, 2016. A new feature was the availablity of
endpoint APIs for accessing data programatically. This tutorial will show
how to do that via R, using the httr package. Additionally, we'll compare that
data with what's found in our BigQuery tables.

The GDC endpoint information is found at:

https://gdc-docs.nci.nih.gov/API/Users_Guide/Getting_Started/#api-endpoints

## First let's check that the system is online.
```{r message=FALSE}
library(httr)
b <- "https://gdc-api.nci.nih.gov/status"
content(GET(url=b))
```

## Finding files at the GDC
The https://gdc-api.nci.nih.gov/cases endpoint enables search and retrieval of
information related to a specific case, or sample donor.

```{r}
b <- "https://gdc-api.nci.nih.gov/cases"
q <- list(cases.samples.portions.analytes.aliquots.aliquot_id="TCGA-DX-A3UC-01A-11R-A30C-07",pretty="true")
res0 <- (GET(url=b, query=q))
resParse <- xmlParse(content(res0, "text"))
```


```{r}
library(XML)

b <- "https://gdc-api.nci.nih.gov/cases"
q <- list(submitter_id="TCGA-DX-A3UC-01A-11R-A30C-07")
res0 <- (GET(url=b, query=q))
resParse <- xmlParse(content(res0, "text"))
resParse
```

We can also retrieve the information as a tab separated table,
although, it's a bit hard to parse/understand.

```{r}
b <- "https://gdc-api.nci.nih.gov/cases"
q <- list(submitter_id="TCGA-BH-A0EA", format="TSV")
res0 <- (GET(url=b, query=q))
res0
headers(res0)
content(res0, as="text")
resText <- content(res0, as="text")
resTable <- read.table(text=resText, sep="\t", header=T, stringsAsFactors=F)

```
