# Working with barcode lists

As part of working with the ISB-CGC web app, you will have created cohorts, represented by lists
of barcodes. This short tutorial shows how to retrieve your cohorts, and even create new ones,
from within the R environment.

The isb-cgc project has a collection of web services called 'endpoints', which accept
and return information. The endpoints allow the user to interact with the isb-cgc system programmatically,
or for a client application to do so on behalf of the user.

The ISBCGCExamples package contains a number of "wrapper" functions, that make calling
the endpoints possible from the R environment. These are just examples, and much
more is possible. See the documentation to get more ideas:

http://isb-cancer-genomics-cloud.readthedocs.io/en/latest/sections/progapi/Programmatic-API.html#isb-cgc-api

To get started, we load up R, and import the ISBCGCExamples library.

```{r}
library(ISBCGCExamples)
```

### Creating a token

The first step is creating a token. This token contains your authentication
status, and lets the service know about what information is available to you.

```{r eval=F}
my_token <- isb_init()
```

Calling the isb_init function is going to open a browser window that lets you
authenticate with Google.
